Thank you, next: demographic consequences of partner diversity and turnover in a multi-species ant-plant mutualism

### 2024-10-12 Manuscript/ant_cactus_AmNat.tex
Manuscript Updates -- Discussion

	\section*{Abstract}
% 200 words max
 The diversity of partners in a multi-species mutualism causes varied demographic effects on the population of the focal mutualist which can be explained by several mechanisms: portfolio effect, complementarity, and sampling effect.
\textit{Cylindropuntia imbricata} (tree cholla) produce extrafloral nectar and various ant partners provide defense from herbivores and seed predators. 
We used plant demographic censuses to parameterize a series of Bayesian hierarchical generalized linear vital rate models to determine the impacts of different partners on the focal mutualists. 
We constructed an Integral Projection Model in which we simulate different combinations of ant partners that don’t occur in nature.
The hierarchical models revealed that different ant partners had different impacts on the cholla vital rates. 
Specifically, \textit{C. opuntiae} tended plants had advantages in both growth and survival when small, and large \textit{L. apiculatum} tended plants had floral viability advantages. 
The IPM results revealed that despite these differences in vital rates, the presence of any ant partner is beneficial, the identity and number of partners do not matter.
This suggests that there are no benefits of diversity in this system.
This study highlights that partner diversity does not always increase the benefits a focal mutualist receives.

\newpage{}
\section*{Introduction}
Mutualisms are species interactions where all participants receive net benefits, leading to higher individual fitness and increased population growth rates. 
They are widespread species interactions but can deteriorate into commensalism or parasitism under conditions that elevate costs or dampen benefits \citep{Rodriguez-Rodriguez2017,Song2020,Mandyam2014,Thrall2007, Bahia2022,Bronstein1994,Chamberlain2014,Frederickson2013,Axelrod1981,Leigh2010}.
Mutualisms are considered more context dependent than other species interactions \citep{Chamberlain2014,Frederickson2013}, meaning the magnitude and sign of interaction strength are often determined by environmental conditions and species' identities \citep{Noe1994,Leigh2010}.

@@

There are multiple mechanisms by which partner diversity can influence the net benefits accrued by a focal mutualist, mirroring the mechanisms by which, at a larger scale of organization, biodiversity can influence ecosystem function \cite{Yeung2006,Barrett2015,Ushio2020}. 
First, when there is a hierarchy of fitness effects (a consistent ranking of best to worst mutualists), a more diverse sample of the partner community may be more likely to include the best partner \cite{Frederickson2013}.
This can lead to an apparent benefit of diversity driven by a sampling effect \cite{Batstone2018} -- but there is no benefit of diversity \emph{per se}, only better and worse partners. 
If partner associations are mutually exclusive (a focal mutualist may engage with only one partner at a time), then partner diversity may impose opportunity costs, leading to negative effects of a diverse mutualist assemblage relative to exclusive association with the single best partner \citep{Miller2007}. 
Second, even within a single mutualist guild, the benefits conferred by alternative partner species can vary in type and not just degree \cite{Stachowicz2005,Bronstein2006,Stanton2003}. 
This can lead to a positive effect of partner diversity through complementarity of alternative functions \cite{Batstone2018}. 
Interference or synergies between partners can make their combined effect different than the expected from the sum of complementary functions \cite{Afkhami2014}. 
Third, partner species can have species-specific responses to environmental variation, either spatially \citep{Ollerton2006} or temporally \citep{Alarcon2008}. 
Multiple partners can therefore act as a 'portfolio' that stabilizes fitness benefits across spatial or temporal heterogeneity, leading to positive effects of partner diversity through the portfolio effect \cite{Batstone2018,Lazaro2022,Horvitz1990}.

Partner diversity can have different effects depending on whether partners are present simultaneously or sequentially (partner turnover) \citep{Djieto-Lordon2005, Ness2006, Bruna2014,Barrett2015,Ushio2020,Dattilo2014}. 
Sequential associations are likely when alternative partners engage in interference competition for access to a shared mutualist \cite{Kiers2003,Batstone2018,Tgaard2015,Wulff2008}. 
Turnover can happen at different timescales, from minutes to years \citep{Oliveira1999,Horvitz1986}. 
The frequency of partner turnover can impact the level of benefits received by the focal mutualist, particularly if the benefits continue to accumulate with successive turnover (e.g., when sequential partners provide complementary functions) or if they saturate over time \citep{Sachs2004,Fiala1994}.
Directionality of turnover can also influence effects of partner diversity if partner identity changes consistently across ontogeny of a focal mutualist \citep{Fonseca2003,Noe1994,Dejean2008}.
For example, plant susceptibility to enemies can change across life stages \citep{Boege2005,Barton2010}, so the benefits of a diverse guild of defensive mutualists are greatest when more defensive partner species align with more vulnerable life stages \citep{Djieto-Lordon2005,Dejean2008}.

Defensive ant-plant mutualisms -- where plants provide food and/or housing to ants that in turn defend them from enemies -- are widespread interactions that offer valuable model systems for the ecology and evolution of mutualism \citep{Bronstein1998, Bronstein2006}. 
Extrafloral nectar (EFN) -bearing plants can serve as dietary resources that promote ant abundance and colony size \citep{Byk2011, Ness2009, Ness2006, Donald2022}.
In turn, presence of defensive ant partners is often linked to reductions in herbivory  \citep{Trager2010, Rudgers2004} and demographic advantages for the plant partner \citep{Baez2016}.
Defensive ant-plant mutualisms are commonly multi-species, where a guild of ant partner species share, and often compete for, a plant mutualist \citep{Bronstein1998, Beattie1985, Trager2010, Agrawal1998}.
Ant partners can vary in their ability to deter herbivores \citep{Bruna2014}, and visitation by low quality ant partners can prevent visitation by higher quality partners, consequently causing a reduction in fitness through missed opportunity costs \citep{Fraser2001, Frederickson2005}.
Susceptibility to herbivory can also vary significantly throughout the life stages of the plant \citep{Boege2005}, suggesting that the order and timing of successive partners is important to the fitness impacts of the combined partner guild \citep{Barton2010, Boege2005, Fonseca2003}.
Finally, herbivore identity and pressure can vary inter-annually \cite{Wetzel2023}, much like mutualist identity and presence, meaning the threat plants face can vary just as much as the protection they receive due to temporal stochasticity. 
Previous studies have investigated how ant partner diversity affects plant fitness \citep{Palmer2010,Afkhami2014,Fiala1994,Gaume1998,Dattilo2014,Ludka2015}
However, little is known about the combined effects of partner identity, directional partner turnover, and temporal stochasticity, particularly because the necessary long-term data are rarely available.

This study examined the consequences of partner diversity in a food-for-protection mutualism between the tree cholla cactus (\textit{Cylindropuntia imbricata}), a long-lived EFN-bearing plant, and multiple species of ant partners.
Previous studies have shown that herbivory by specialized insect herbivores negatively affects cactus fitness \cite{Miller2009}, and that ant defense reduces herbivore damage \cite{Miller2007}. 
Tree cholla are tended by two common ant species (\textit{Liometopum apiculatum} and \textit{Crematogaster opuntiae}) and several additional rarer species, all of which collect EFN during foraging visits but their colonies are ground-nesting and not housed by the plants. 
These ant species locally co-occur but individual plants are typically tended by only one species that patrols the plant around-the-clock and maintains control of the plant's nectar resources, usually for an entire growing season \citep{Ohm2014, Donald2022}. 
Switches between partner species, or between vacancy and ant occupancy, commonly occur from one growing season to the next \citep{Miller2007}. 
Prior experiments suggested a hierarchy of mutualist quality, with \textit{L. apiculatum} providing strong anti-herbivore defense and \textit{C. opuntiae} having net negative effects because herbivore deterrence is outweighed by deterrence of pollinators \citep{Miller2007,Ohm2014}. 
However, all of our previous studies in this system have focused on single life stages (adult plants) or vital rates (seed production) and did not integrate the demographic effects of ant defense across the life cycle, which may be essential for understanding net fitness effects of a diverse partner guild \citep[e.g.,][]{Palmer2010}. 
To our knowledge no previous study has incorporated inter-annual stochasticity into models of ant-plant dynamics, which limits our understanding of diversity benefits that may arise through the portfolio effect.

Here we used a unique long-term data set that allows us to explore mutualistic associations with multiple partner species, longitudinal turnover in partner identity, and how the demographic effects of alternative partners varies across plant size structure and nearly 20 years of inter-annual fluctuations. 
We used this observational data set of plant demography and ant-plant associations, contextualized by previous ant exclusion experiments, to investigate whether and through which mechanism(s) partner diversity affects the fitness benefits of ant visitation. 
Specifically, we asked:
	\begin{enumerate}	
		\item{What are the demographic effects of association with alternative partners and how do these effects fluctuate across years?}
		\item{What are the frequency and direction of partner turnover across the plant life cycle?}	
		\item{What is the net effect of partner diversity on plant fitness, and what mechanism(s) explain(s) this effect?}
	\end{enumerate}
To answer these questions we used a hierarchical Bayesian statistical approach to estimate demographic vital rates for hosts in different states of ant occupancy and to quantify state-dependent partner turnover from the long-term data. 
We then used a stochastic, multi-state integral projection model (IPM) that combines diverse effects on vital rates and pathways of partner turnover to quantify effects of partner diversity on plant fitness.

@@

\noindent %In these equations $x$ and $x'$ indicate the size of a plant in years $t$ and $t+1$ respectively, $a$ and $a'$ indicate the ant partner of a plant in years $t$ and $t+1$.
Functions $P(x;\pmb{\tau}^P)$ and $F(x;\pmb{\tau}^F)$ give the probability of flowering in year $t$ and the number of flowerbuds produced in year $t$, respectively, by plants of size $x$ in year $t$. 
The proportion of flowerbuds that remain viable through fruit set ($V(a;\pmb{\tau}^V_{a})$) and the number of seeds per fruit ($\kappa(a')$) is dependent on ant state $a$. 
The vectors $\pmb{\tau}$ give year-specific deviates (mean zero) and appear in functions for which we can estimate temporal stochasticity from the long-term data; superscripts indicate the corresponding vital rate and, when present, the $a$ subscript indicates that deviates are specific to plants in ant state $a$.
For example, temporal deviates $\pmb{\tau}^V_{a}$ describe better- and worse-than-average years for flowerbud viability and plants in different ant states can fluctuate independently (good years for \textit{L. apiculatum} -occupied plants may not be good years for \textit{C. opuntiae}-occupied plants, for example). 
Seed production is integrated over the size distribution, from the lower $L$ to upper $U$ size limits, and summed over all possible initial ant states ($A=4$) giving total seed production. 
Seeds are multiplied by the probability of escaping post-dispersal seed predation ($\delta$) to give the number of seeds that enter the one-year old seed bank. 
Plants can recruit out of the year-one seed bank with probability $\gamma_1$ or transition to the two-year seed bank with a probability of $1 - \gamma_1$. 
Seeds in the two-year seed bank are assumed to either germinate with probability $\gamma_2$ or die.

@@

\paragraph{Growth}
We fit the growth sub-model ($G(x',x,a;\pmb{\tau^G_{a}})$) to data on size in year $t+1$ ($y^G$) using the skewed normal distribution to account for left-skewed size transitions (at some initial sizes, transitions below the expected future size were more common than transitions above it). 
The skew-normal has three parameters corresponding to location ($\hat{G}$), shape ($\sigma$), and scale ($\alpha$):
\begin{linenomath*}
	$$y_i^G \sim Skewed Normal(\hat{G_i},\sigma_i,\alpha_i) $$
	$$\hat{G_i} = \beta^0_{a[i]} + \beta^1_{a[i]} x_i + \beta^2_{a[i]} x_i^2 + u_{year[i],a[i]} + w_{plot[i]} $$
	$$log(\sigma_i)  = \beta^3 + \beta^4 x_i $$
	$$\alpha_i = \beta^5 + \beta^6 x_i$$
	\label{eqn:growth}
\end{linenomath*}
Here, the location parameter for the $i$th observation $\hat{G_i}$ is defined as a second-order polynomial with ant-size interactions because  preliminary analysis found this was an improvement over a linear relationship (note that the location parameter of the skew-normal is not the mean, but the mean can be derived as $\hat{G} + \frac{\sigma\alpha}{\sqrt{1+\alpha^2}} \sqrt{\frac{2}{\pi}}$). 
The year- and ant-specific random effect $u$ (which parameterizes the $\pmb{\tau}^G_{a}$ vectors) and plot-specific random effect $w$ are normally distributed with variances $\sigma^2_{year}$ and $\sigma^2_{plot}$, respectively. 
Parameters $\sigma_i$ and $\alpha_i$  control residual variance and skewness, respectively, and were defined as linear functions of initial size $x_i$ ($\sigma_i$ is strictly positive and was modeled with a log link function). 
We assume growth variance and skewness were not dependent on ant occupancy state.

@@

The flowerbud viability sub-model ($V(a;\pmb{\tau}^{V}_{a})$) estimates the probability that flowerbuds produced in year $t$ remain viable (not aborted), with fixed effects of ant partner $a$ in year $t$.
We fit this model to floral viability data using a binomial distribution where trials and successes are given by the total number of flowerbuds and the number that are viable, respectively.
This model used a logit link function and included random effects for plot and year following the same structure as the growth and survival models, with ant-specific year random effects.

@@

\paragraph{Parameter estimation}
We fit models using Stan run through version 4.0.2 of R \cite{Rcite,Rstancite}. 
We used vague priors for all parameters. 
For each model, we obtained three chains of 10,000 iterations, discarding the first 1,500 iterations. 
We visually assessed parameter convergence between and within chains (Appendix \ref{appendix:D}: Figures \ref{fig:Surv_post} -- \ref{fig:Pre_Surv_post} b) and assessed overall model fit with posterior predictive checks to examine how well the fitted model can generate simulated data similar to the real data (Appendix \ref{appendix:D}: Figures \ref{fig:Surv_post} -- \ref{fig:Pre_Surv_post} a).

\subsection*{IPM Analysis}
Analyzing the IPM required that we discretize the continuous IPM kernel into an approximating matrix. 
Size variable $x$ was discretized into $b$ bins, resulting in a $b \times b$ matrix.
In our model there is additional complexity in the form of transitions between $A$ ant states and two additional discrete states (year one and year two seed banks), leading to a matrix size of $A(b+2) \times A(b+2)$.
We used $b = 500$ bins, which we found to be sufficient for numerically stable outputs, and extended the integration limits beyond the minimum ($L$) and maximum ($U$) observed sizes to avoid unintentional eviction using the ``floor-and-ceiling'' method \cite{Williams2012}.

For stochastic analyses, we estimated the approximating matrix corresponding to each $t$ to $t+1$ transition year. 
To estimate population mean fitness in a stochastic environment ($\lambda_{S}$) we simulated population dynamics for 500 years by randomly sampling among the 15 annual transition matrices, discarding the first 100 years of the simulation to minimize the influence of initial conditions. 
Sampling observed transition matrices (rather than independently sampling regression coefficients) produces demographic time series that realistically capture inter-annual variation by preserving correlations between vital rates \cite{metcalf2015statistical}.
We tallied the total population size at each time step as  $N_{t} = B^1_{t} + B^2_{t} + \sum_{a=1}^{A}\int n(x,a)_{t}dx$ and calculated the stochastic growth rate as $log(\lambda_S) = E[log(\frac{N_{t+1}}{N_{t}})]$ \citep{Mark2009}.
We propagated uncertainty from the vital rate models using 100 draws from the joint posterior distribution of model parameters, resulting in a posterior distribution of $\lambda_{S}$ and other derived quantities.

@@

The partner diversity treatment scenarios required additional assumptions about the mechanisms that give rise to observed occupancy patterns. 
Based on evidence that EFN-bearing cacti are nearly always ant-occupied \citep{Miller2014}, we assume that ant partners competitively exclude one another from EFN-bearing cacti and that competition is zero-sum. 
This means that, in scenarios that remove species from the partner community, remaining species gain access to plants that the removed species would have tended. 
In Appendix \ref{appendix}, we present results under an alternative assumption, that ant visitation is limited by factors other than availability of cactus EFN (e.g., nesting sites or off-plant dietary resources), such that when a species is removed from the partner community, the plants it would have occupied remain vacant.

@@

\section*{Results}
\subsection*{What are the demographic effects of association with alternative partners and how do these effects fluctuate across years?}
Over the 20-year data set, we found that ant partners influenced demographic performance of cactus hosts, and different ant partners had contrasting demographic effects across host vital rates. 
Ant-tended plants had a growth advantage over vacant plants, especially at smaller sizes (Figure \ref{fig:Grow}).
Over the size range for which plants were likely ant-tended (larger than the minimum observed size of $0.8$ $log(cm^3)$; solid lines in Figure \ref{fig:Grow}) there were modest differences between partner species, with the greatest growth advantage associated with \textit{C. opuntiae} followed by \textit{L. apiculatum} and then Other ants.
At large sizes the growth trajectories of ant-tended and vacant plants were nearly indistinguishable. 
For all ant states, growth was left-skewed, with transitions to sizes below the mean were more common than sizes above the mean.

@@

\begin{figure}
	\includegraphics[width = 0.95\linewidth]{Figures/grow_contour_v2.png}
	\caption{This figure shows the next predicted size of cholla based on previous size with each individual ant partner. The colored lines (seen in all panels) are the next mean predicted size of cholla. Dashed lines indicate extrapolation while solid lines indicate estimations within the range of observed data. The points (seen in panels a-d) are the observed data which informs these estimates. The black contour lines (seen in a-d) appear at 5\% increments showing where 5\%, 10\%, etc. of the data is expected to fall. They gray dashed line (in panel e only) shows the line where the next predicted size is the same as the previous (aka there is no growth on this line and below this line is shrinkage).}
	\label{fig:Grow}
\end{figure}

>>>>

	\section*{Abstract}
% 200 words max
 The diversity of partners in a multi-species mutualism causes varied demographic effects on the population of the focal mutualist which can be explained by several mechanisms: portfolio effect, complementarity, and sampling effect.
 To study multi-species demographic effects, we considered \textit{Cylindropuntia imbricata} (tree cholla), which produce extrafloral nectar and various ant partners ( \textit{Crematogaster opuntiae}, \textit{Liometopum apiculatum}, and other less frequent ants) provide defense from herbivores and seed predators. 
We used plant demographic censuses to parameterize a series of Bayesian hierarchical generalized linear vital rate models to determine the impacts of different partners on the focal mutualists. 
We constructed an Integral Projection Model (IPM) in which we simulate different combinations of ant partners that don’t occur in nature.
Different ant partners had different impacts on the cholla vital rates. 
Specifically, \textit{C. opuntiae} tended plants had advantages in both growth and survival when small, and large \textit{L. apiculatum} tended plants had floral viability advantages. 
The IPM results revealed that despite these differences in vital rates, the presence of any ant partner is beneficial, the identity and number of partners do not matter.
This suggests that there are no benefits of diversity in this system and therefore than none of the highlighted mechanisms explain the benefits of multi-species mutualisms.
This study highlights that partner diversity does not always increase the benefits a focal mutualist receives.

\newpage{}
\section*{Introduction}
Cooperations between species, aka mutualisms, occur on a pairwise basis or in a more complex network and often play integral and foundational roles in ecological communities.
Mutualisms are species interactions where all participants receive net benefits, leading to higher individual fitness and increased population growth rates. 
They are widespread species interactions but can deteriorate into commensalism or parasitism under conditions that elevate costs or dampen benefits \citep{Rodriguez-Rodriguez2017,Song2020,Mandyam2014,Thrall2007, Bahia2022,Bronstein1994,Chamberlain2014,Frederickson2013,Axelrod1981,Leigh2010}.
Mutualisms are considered more context dependent than other species interactions \citep{Chamberlain2014,Frederickson2013}, meaning the magnitude and sign of interaction strength are often determined by environmental conditions and species' identities \citep{Noe1994,Leigh2010}.

@@

There are multiple mechanisms by which partner diversity can influence the net benefits accrued by a focal mutualist, mirroring the mechanisms by which, at a larger scale of organization, biodiversity can influence ecosystem function \citep{Yeung2006,Barrett2015,Ushio2020}. 
First, when there is a hierarchy of fitness effects (a consistent ranking of best to worst mutualists), a more diverse sample of the partner community may be more likely to include the best partner \citep{Frederickson2013}.
This can lead to an apparent benefit of diversity driven by a sampling effect \citep{Batstone2018} -- but there is no benefit of diversity \emph{per se}, only better and worse partners. 
If partner associations are mutually exclusive (a focal mutualist may engage with only one partner at a time), then partner diversity may impose opportunity costs, leading to negative effects of a diverse mutualist assemblage relative to exclusive association with the single best partner \citep{Miller2007}. 
Second, even within a single mutualist guild, the benefits conferred by alternative partner species can vary in type and not just degree \citep{Stachowicz2005,Bronstein2006,Stanton2003}. 
This can lead to a positive effect of partner diversity through complementarity of alternative functions \citep{Batstone2018}. 
Interference or synergies between partners can make their combined effect different than the expected from the sum of complementary functions \citep{Afkhami2014}. 
Third, partner species can have species-specific responses to environmental variation, either spatially \citep{Ollerton2006} or temporally \citep{Alarcon2008}. 
Multiple partners can therefore act as a 'portfolio' that stabilizes fitness benefits across spatial or temporal heterogeneity, leading to positive effects of partner diversity through the portfolio effect \citep{Batstone2018,Lazaro2022,Horvitz1990}.

Partner diversity can have different effects depending on whether partners are present simultaneously or sequentially (partner turnover) \citep{Djieto-Lordon2005, Ness2006, Bruna2014,Barrett2015,Ushio2020,Dattilo2014}. 
Sequential associations are likely when alternative partners engage in interference competition for access to a shared mutualist \citep{Kiers2003,Batstone2018,Tgaard2015,Wulff2008}. 
Turnover can happen at different timescales, from minutes to years \citep{Oliveira1999,Horvitz1986}. 
The frequency of partner turnover can impact the level of benefits received by the focal mutualist, particularly if the benefits continue to accumulate with successive turnover (e.g., when sequential partners provide complementary functions) or if they saturate over time \citep{Sachs2004,Fiala1994}.
Directionality of turnover can also influence effects of partner diversity if partner identity changes consistently across ontogeny of a focal mutualist \citep{Fonseca2003,Noe1994,Dejean2008}.
For example, plant susceptibility to enemies can change across life stages \citep{Boege2005,Barton2010}, so the benefits of a diverse guild of defensive mutualists are greatest when more defensive partner species align with more vulnerable life stages \citep{Djieto-Lordon2005,Dejean2008}.

Defensive ant-plant mutualisms -- where plants provide food and/or housing to ants that in turn defend them from enemies -- are widespread interactions that offer valuable model systems for the ecology and evolution of mutualism \citep{Bronstein1998, Bronstein2006}. 
Extrafloral nectar (EFN) -bearing plants can serve as dietary resources that promote ant abundance and colony size \citep{Byk2011, Ness2009, Ness2006, Donald2022}.
In turn, presence of defensive ant partners is often linked to reductions in herbivory  \citep{Trager2010, Rudgers2004} and demographic advantages for the plant partner \citep{Baez2016}.
Defensive ant-plant mutualisms are commonly multi-species, where a guild of ant partner species share, and often compete for, a plant mutualist \citep{Bronstein1998, Beattie1985, Trager2010, Agrawal1998}.
Ant partners can vary in their ability to deter herbivores \citep{Bruna2014}, and visitation by low quality ant partners can prevent visitation by higher quality partners, consequently causing a reduction in fitness through missed opportunity costs \citep{Fraser2001, Frederickson2005}.
Susceptibility to herbivory can also vary significantly throughout the life stages of the plant \citep{Boege2005}, suggesting that the order and timing of successive partners is important to the fitness impacts of the combined partner guild \citep{Barton2010, Boege2005, Fonseca2003}.
Finally, herbivore identity and pressure can vary inter-annually \citep{Wetzel2023}, much like mutualist identity and presence, meaning the threat plants face can vary just as much as the protection they receive due to temporal stochasticity. 
Previous studies have investigated how ant partner diversity affects plant fitness \citep{Palmer2010,Afkhami2014,Fiala1994,Gaume1998,Dattilo2014,Ludka2015}
However, little is known about the combined effects of partner identity, directional partner turnover, and temporal stochasticity, particularly because the necessary long-term data are rarely available.

This study examined the consequences of partner diversity in a food-for-protection mutualism between the tree cholla cactus (\textit{Cylindropuntia imbricata}), a long-lived EFN-bearing plant, and multiple species of ant partners.
Previous studies have shown that herbivory by specialized insect herbivores negatively affects cactus fitness \citep{Miller2009}, and that ant defense reduces herbivore damage \citep{Miller2007}. 
Tree cholla are tended by two common ant species (\textit{Liometopum apiculatum} and \textit{Crematogaster opuntiae}) and several additional rarer species, all of which collect EFN during foraging visits but their colonies are ground-nesting and not housed by the plants. 
These ant species locally co-occur but individual plants are typically tended by only one species that patrols the plant around-the-clock and maintains control of the plant's nectar resources, usually for an entire growing season \citep{Ohm2014, Donald2022}. 
Switches between partner species, or between vacancy and ant occupancy, commonly occur from one growing season to the next \citep{Miller2007}. 
Prior experiments suggested a hierarchy of mutualist quality, with \textit{L. apiculatum} providing strong anti-herbivore defense and \textit{C. opuntiae} having net negative effects because herbivore deterrence is outweighed by deterrence of pollinators \citep{Miller2007,Ohm2014}. 
However, all of our previous studies in this system have focused on single life stages (adult plants) or vital rates (seed production) and did not integrate the demographic effects of ant defense across the life cycle, which may be essential for understanding net fitness effects of a diverse partner guild \citep[e.g.,][]{Palmer2010}. 
To our knowledge no previous study has incorporated inter-annual stochasticity into models of ant-plant dynamics, which limits our understanding of diversity benefits that may arise through the portfolio effect.

Here we used a unique long-term data set that allows us to explore mutualistic associations with multiple partner species, longitudinal turnover in partner identity, and how the demographic effects of alternative partners varies across plant size structure and nearly 20 years of inter-annual fluctuations. 
We used this observational data set of plant demography and ant-plant associations, contextualized by previous ant exclusion experiments, to investigate whether and through which mechanism(s) partner diversity affects the fitness benefits of ant visitation. 
Specifically, we asked:
	\begin{enumerate}	
		\item{What are the vital rate effects of association with alternative partners and how do these effects fluctuate across years?}
		\item{What are the frequency and direction of partner turnover across the plant life cycle?}	
		\item{What is the net effect of partner diversity on plant fitness, and what mechanism(s) explain(s) this effect?}
	\end{enumerate}
To answer these questions we used a hierarchical Bayesian statistical approach to estimate demographic vital rates for hosts in different states of ant occupancy and to quantify state-dependent partner turnover from the long-term data. 
We then used a stochastic, multi-state integral projection model (IPM) that combines diverse effects on vital rates and pathways of partner turnover to quantify effects of partner diversity on plant fitness.

@@

\noindent %In these equations $x$ and $x'$ indicate the size of a plant in years $t$ and $t+1$ respectively, $a$ and $a'$ indicate the ant partner of a plant in years $t$ and $t+1$.
Functions $P(x;\pmb{\tau}^P)$ and $F(x;\pmb{\tau}^F)$ give the probability of flowering in year $t$ and the number of flowerbuds produced in year $t$, respectively, by plants of size $x$ in year $t$. 
The proportion of flowerbuds that remain viable through fruit set ($V(a;\pmb{\tau}^V_{a})$) and the number of seeds per fruit ($\kappa(a)$) is dependent on ant state $a$. 
The vectors $\pmb{\tau}$ give year-specific deviates (mean zero) and appear in functions for which we can estimate temporal stochasticity from the long-term data; superscripts indicate the corresponding vital rate and, when present, the $a$ subscript indicates that deviates are specific to plants in ant state $a$.
For example, temporal deviates $\pmb{\tau}^V_{a}$ describe better- and worse-than-average years for flowerbud viability and plants in different ant states can fluctuate independently (good years for \textit{L. apiculatum} -occupied plants may not be good years for \textit{C. opuntiae}-occupied plants, for example). 
Seed production is integrated over the size distribution, from the lower $L$ to upper $U$ size limits, and summed over all possible initial ant states ($A=4$) giving total seed production. 
Seeds are multiplied by the probability of escaping post-dispersal seed predation ($\delta$) to give the number of seeds that enter the one-year old seed bank. 
Plants can recruit out of the year-one seed bank with probability $\gamma_1$ or transition to the two-year seed bank with a probability of $1 - \gamma_1$. 
Seeds in the two-year seed bank are assumed to either germinate with probability $\gamma_2$ or die.

@@

\paragraph{Growth}
We fit the growth sub-model ($G(x',x,a;\pmb{\tau^G_{a}})$) to data on size in year $t+1$ ($y^G$) using the skewed normal distribution to account for left-skewed size transitions (at some initial sizes, transitions below the expected future size were more common than transitions above it). 
The skew-normal has three parameters corresponding to location ($\hat{G}$), shape ($\sigma$), and scale ($\alpha$):
\begin{linenomath*}
	$$y_i^G \sim Skewed Normal(\hat{G_i},\sigma_i,\alpha_i) $$
	$$\hat{G_i} = \beta^0_{a[i]} + \beta^1_{a[i]} x_i + \beta^2_{a[i]} x_i^2 + u_{year[i],a[i]} + w_{plot[i]} $$
	$$log(\sigma_i)  = \beta^3 + \beta^4 x_i $$
	$$\alpha_i = \beta^5 + \beta^6 x_i$$
	\label{eqn:growth}
\end{linenomath*}
Here, the location parameter for the $i$th observation $\hat{G_i}$ is defined as a second-order polynomial with ant-size interactions because  preliminary analysis found this was an improvement over a linear relationship.
The location parameter of the skew-normal is not the mean, but the mean can be derived as $\hat{G} + \frac{\sigma\alpha}{\sqrt{1+\alpha^2}} \sqrt{\frac{2}{\pi}}$. 
The year- and ant-specific random effect $u$ (which parameterizes the $\pmb{\tau}^G_{a}$ vectors) and plot-specific random effect $w$ are normally distributed with variances $\sigma^2_{year}$ and $\sigma^2_{plot}$, respectively. 
Parameters $\sigma_i$ and $\alpha_i$  control residual variance and skewness, respectively, and were defined as linear functions of initial size $x_i$ ($\sigma_i$ is strictly positive and was modeled with a log link function). 
We assume growth variance and skewness were not dependent on ant occupancy state.

@@

The flowerbud viability sub-model ($V(a;\pmb{\tau}^{V}_{a})$) estimates the probability that flowerbuds produced in year $t$ remain viable (not aborted), with fixed effects of ant partner $a$ in year $t$.
We fit this model to floral viability data using a binomial distribution where trials and successes are given by the total number of flowerbuds and the number that are viable, respectively.
This model used a logit link function and included random effects for plot and year following the same structure as the survival model, with ant-specific year random effects.

@@

\paragraph{Parameter estimation}
We fit models using Stan run through version 4.0.2 of R \citep{Rcite,Rstancite}. 
We used vague priors for all parameters. 
For each model, we obtained three chains of 10,000 iterations, discarding the first 1,500 iterations. 
We visually assessed parameter convergence between and within chains (Figures S\ref{fig:Surv_post} -- \ref{fig:Pre_Surv_post}b) and assessed overall model fit with posterior predictive checks to examine how well the fitted model can generate simulated data similar to the real data (Figures S\ref{fig:Surv_post} -- \ref{fig:Pre_Surv_post}a).

\subsection*{IPM Analysis}
Analyzing the IPM required that we discretize the continuous IPM kernel into an approximating matrix. 
Size variable $x$ was discretized into $b$ bins, resulting in a $b \times b$ matrix.
In our model there is additional complexity in the form of transitions between $A$ ant states and two additional discrete states (year one and year two seed banks), leading to a matrix size of $A(b+2) \times A(b+2)$.
We used $b = 500$ bins, which we found to be sufficient for numerically stable outputs, and extended the integration limits beyond the minimum ($L$) and maximum ($U$) observed sizes to avoid unintentional eviction using the ``floor-and-ceiling'' method \citep{Williams2012}.

For stochastic analyses, we estimated the approximating matrix corresponding to each $t$ to $t+1$ transition year. 
To estimate population mean fitness in a stochastic environment ($\lambda_{S}$) we simulated population dynamics for 500 years by randomly sampling among the 15 annual transition matrices, discarding the first 100 years of the simulation to minimize the influence of initial conditions. 
Sampling observed transition matrices (rather than independently sampling regression coefficients) produces demographic time series that realistically capture inter-annual variation by preserving correlations between vital rates \citep{metcalf2015statistical}.
We tallied the total population size at each time step as  $N_{t} = B^1_{t} + B^2_{t} + \sum_{a=1}^{A}\int n(x,a)_{t}dx$ and calculated the stochastic growth rate as 
$$log(\lambda_S) = E[log(\frac{N_{t+1}}{N_{t}})]$$ \citep{Mark2009}.
We propagated uncertainty from the vital rate models using 100 draws from the joint posterior distribution of model parameters, resulting in a posterior distribution of $\lambda_{S}$ and other derived quantities.

@@

The partner diversity treatment scenarios required additional assumptions about the mechanisms that give rise to observed occupancy patterns. 
Based on evidence that EFN-bearing cacti are nearly always ant-occupied \citep{Miller2014}, we assume that ant partners competitively exclude one another from EFN-bearing cacti and that competition is zero-sum. 
This means that, in scenarios that remove species from the partner community, remaining species gain access to plants that the removed species would have tended. 
In Appendix C, we present results under an alternative assumption, that ant visitation is limited by factors other than availability of cactus EFN (e.g., nesting sites or off-plant dietary resources), such that when a species is removed from the partner community, the plants it would have occupied remain vacant.

@@

\section*{Results}
\subsection*{What are the effects of association with alternative partners on vital rates and how do these effects fluctuate across years?}
Over the 20-year data set, we found that ant partners influenced demographic performance of cactus hosts, and different ant partners had contrasting demographic effects across host vital rates. 
Ant-tended plants had a growth advantage over vacant plants, especially at smaller sizes (Figure \ref{fig:Grow}).
Over the size range for which plants were likely ant-tended (larger than the minimum observed size of $0.8$ $log(cm^3)$; solid lines in Figure \ref{fig:Grow}) there were modest differences between partner species, with the greatest growth advantage associated with \textit{C. opuntiae} followed by \textit{L. apiculatum} and then Other ants.
At large sizes the growth trajectories of ant-tended and vacant plants were nearly indistinguishable. 
For all ant states, growth was left-skewed, with transitions to sizes below the mean were more common than sizes above the mean.

@@

\begin{figure}
	\includegraphics[width = 0.95\linewidth]{Figures/grow_contour_v2.png}
	\caption{The next predicted size of cholla based on previous size with each individual ant partner. The colored lines (seen in all panels) are the next mean predicted size of cholla. Dashed lines indicate extrapolation while solid lines indicate estimations within the range of observed data. The points (seen in panels a-d) are the observed data which informs these estimates. The black contour lines (seen in a-d) appear at 5\% increments showing where 5\%, 10\%, etc. of the data is expected to fall. Panel e shows all mean estimates from panels a-d on one graph. The gray dashed line (in panel e only) shows the line where the next predicted size is the same as the previous (aka there is no growth on this line and below this line is shrinkage).}
	\label{fig:Grow}
\end{figure}

\begin{figure}[H]
\includegraphics[width=0.95\linewidth]{Figures/survival_plot.png}
\caption{The estimated survival rates based on the size of the cactus with each individual ant partner. The solid colored lines (shown on all panels) indicate the mean estimated survival rates. The dashed lines (shown in panel e) indicate extrapolations beyond existing data (where we estimated survival for plants tended by ants where we had never seen a tended cactus of that size). The grey area around the solid lines (shown in panels a-d) show the 90\% confidence interval for the estimates. The colored dots are the real data binned by size to show how our estimates align with real survival observations. A larger circle means we had more data on survival of plants of this size with this partner. The mean estimates from panels a-d are combined on one graph in panel e.}
\label{fig:Surv}
\end{figure}
